Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments

### 2017-01-13 removals/manuscript/removal_ms.tex
Minor edits

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions. The models predict little competitive release following species removals.
\item We conducted a removal experiment to test this prediction. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}. We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. If our ``baseline" model, which accounts for local plant-plant interactions, accurately predicts responses to removal, then the fitted removal indicator effect should be non-signficant.
\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model. For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.

@@

To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.  Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data effectively predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level. During the period of data collection (1926 \textendash 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

@@

All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted. The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.

@@

Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs. removal plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects. We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

@@

Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}. We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure). Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect. If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero. However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant. For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots. This result would indicate that our "baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release. We refer to the model that includes the estimated removal treatment effect as the "removal model."

Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.

@@

Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
We fit the models using the \textit{R-INLA} package for R \citep{rue_approximate_2009}. To explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{Artemisia} shrubs, we fit models using the \textit{lmer} function in the \textit{lme4} package for R

@@

We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters. Each model was run for 20,000 iterations and two independent chains with different initial values for parameters. We discarded the initial 10,000 samples. Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.

For all three vital rate models, we considered the effect of removal treatments to be significant when the 95\% confidence interval (or posterior
interval) on the estimate of $\chi$ did not overlap zero. In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model. For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

@@

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
with a 'no-overlap' rule \citep{adler_coexistence_2010}. In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.

@@

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ. As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero. We did not simulate long-term \textit{A. tripartita} response to grass removal because the vital rate regressions for \textit{A. tripartita} showed that the baseline model effectively predicted removal effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters. If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots. For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots. However, for the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{Poa secunda}  and \textit{P. spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, we could not detect significant removal treatment effects on recruitment.

@@

The positive removal effects on \textit{Poa secunda}  and \textit{P. spicata} growth suggest that models based on observational data may underestimate competitive release. We conducted additional analyses to further explore removal effects on the growth of these two species. First, we classified all individuals of these species as located either inside or outside the limits of \textit{A. tripartita} canopies present in 2011 (before removal). If removal is releasing grasses from \textit{A. tripartita} competition, we would expect individuals located under the shrub canopies to experience a greater effect of removal. However, the location of individuals with respect to \textit{A. tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A. tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.

Second, we evaluated the possibility that removal effects might be transient. We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients. We tested for a transient removal effect by adding a year-by-treatment interaction. For both  \textit{Poa secunda}  and \textit{P. spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).

@@

Finally, we evaluated pre-existing differences in community composition in removal and control plots. In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species. Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A. tripartita}). To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments. \textit{A. tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H. comata} was lower in removal plots than controls, though it was absent from most \textit{P. spicata} neighborhoods in both treatments (Fig. \ref{fig:W-by-treatment}). These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.

@@

To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects. Our models tended to under-predict population growth (Fig. \ref{fig:CoverChange}) and population size (Fig. \ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species. For \textit{A. tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small. \textit{H. comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates. For \textit{Poa secunda} and especially \textit{P. spicata}, the model with removal effects consistently predicted higher population growth than the baseline model  and reduced the underestimates of cover (Fig. \ref{fig:IBM1step}). Viewing all annual growth rates across species and years shows the same pattern of underprediction for some species, but also reveals a strong correlation between observations and predictions in control plots (Fig. \ref{fig:ObsPred1to1}A). In removal plots, underprediction of growth rates in certain years was larger, especially for \textit{H. comata} and \textit{P. spicata}; including removal treatment effects only slightly reduced this bias (Fig. \ref{fig:ObsPred1to1}B). For the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P. secunda} and \textit{P. spicata} (Figs. \ref{fig:IBM1step-maxCI},\ref{fig:ObsPred1to1-maxCI}).

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step}
 \caption{Observed and predicted cover for each of the four modeled species. Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats. Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients. Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
 \label{fig:IBM1step}
 \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{cover_change_1to1}
  \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments. The black line is a 1:1 line. Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects. Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
  \label{fig:ObsPred1to1}
  \end{figure}

>>>>

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions. The models predict little competitive release following species removals.
\item We conducted a removal experiment to test this prediction. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}. We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-signficant.
\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model. For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{A. tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{A. tripartita} removal projected greater competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.

@@

To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.  Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level. During the period of data collection (1926 \textendash 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Hesperostipa comata}, \textit{Pseudoroegneria spicata},  and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

@@

All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted. The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests.git}.

@@

Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs. removal plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \texttt{lme4} package \citep{Bates2015} of R, including year and quadrat as random effects. We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

@@

Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}. We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure). Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect. If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero. However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant. For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots. This result would indicate that our ``baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release. We refer to the model that includes the estimated removal treatment effect as the ``removal model."

Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{H. comata}, \textit{P. spicata}, and \textit{P. secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.

@@

Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach to explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

@@

We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters. Each model was run for 20,000 iterations and two independent chains with different initial values for parameters. We discarded the initial 10,000 samples. Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}. For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% confidence interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.

@@

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
with a `no-overlap' rule \citep{adler_coexistence_2010}. In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.

@@

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ. As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero. We did not simulate long-term \textit{A. tripartita} response to grass removal because the vital rate regressions for \textit{A. tripartita} showed that the baseline model effectively predicted removal effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{P. secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters. If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots. For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots. For the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, we could not detect significant removal treatment effects on recruitment.

@@

The positive removal effects on \textit{P. secunda}  and \textit{P. spicata} growth suggest that models based on observational data may underestimate competitive release. We conducted additional analyses to further explore removal effects on the growth of these two species. First, we classified all individuals of these species as located either inside or outside the limits of \textit{A. tripartita} canopies present in 2011 (before removal). If removal is releasing grasses from \textit{A. tripartita} competition, we would expect individuals located under the shrub canopies to experience a greater effect of removal. However, the location of individuals with respect to \textit{A. tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A. tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.

Second, we evaluated the possibility that removal effects might be transient. We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients. We tested for a transient removal effect by adding a year-by-treatment interaction. For both  \textit{P. secunda}  and \textit{P. spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).

@@

Finally, we evaluated pre-existing differences in community composition in removal and control plots. In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species. Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A. tripartita}). To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments. \textit{A. tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater in removal plots than in the controls, whereas crowding by \textit{H. comata} was lower in removals than controls, though it was absent from most \textit{P. spicata} neighborhoods in both treatments (Fig. \ref{fig:W-by-treatment}). These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.

@@

To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects. Our models tended to under-predict population growth (Fig. \ref{fig:CoverChange}) and population size (Fig. \ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species. For \textit{A. tripartita}, both observed and predicted population growth rates were close to zero, and differences in predictions from the two models were small. \textit{H. comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates. For \textit{Poa secunda} and especially \textit{P. spicata}, the model with removal effects consistently predicted higher population growth than the baseline model  and reduced the underestimates of cover (Fig. \ref{fig:IBM1step}). Viewing all annual growth rates across species and years shows the same pattern of underprediction for some species, but also reveals a strong correlation between observations and predictions in control plots (Fig. \ref{fig:ObsPred1to1}A). In removal plots, underprediction of growth rates in certain years was larger, especially for \textit{H. comata} and \textit{P. spicata}; including removal treatment effects only slightly reduced this bias (Fig. \ref{fig:ObsPred1to1}B). For the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P. secunda} and \textit{P. spicata} (Figs. \ref{fig:IBM1step-maxCI},\ref{fig:ObsPred1to1-maxCI}).

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step}
 \caption{Observed and predicted cover for each of the four modeled species. Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats. Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients. Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
 \label{fig:IBM1step}
 \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{cover_change_1to1}
  \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments. The black line is a 1:1 line. Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects. Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
  \label{fig:ObsPred1to1}
  \end{figure}

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well for predicting \textit{A. tripartita} responses to grass removal). Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2-fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}). Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P. spicata} (Fig. \ref{fig:IPMresults-maxCI}).
